---
title: "Data Analysis Project"
author: "michael purdie, kylie rau, shreya vuttaluru"
date: "12/5/21"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are working with the [Washington Post police shooting database](https://github.com/washingtonpost/data-police-shootings) and the [Washington Post homicides database.](https://github.com/washingtonpost/data-homicides)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
#essential libraries
library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
#potential libraries 
library(janitor)
library(lubridate)

#API Census key
census_api_key("b559c55dc9a86baaa9ab42174d2fa978a618dff1")

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
#Load required data
police_shootings <- read_csv("data/data-police-shootings-master/fatal-police-shootings-data.csv")

#cleaning and organizing
#geocoded data
data1 <- read_rds("data/geocoded_results_500.rds")
data2 <- read_rds("data/geocoded_results_1000.rds")
data3 <- read_rds("data/geocoded_results_1500.rds")
data4 <- read_rds("data/geocoded_results_2000.rds")
data5 <- read_rds("data/geocoded_results_2500.rds")
data6 <- read_rds("data/geocoded_results_3000.rds")
data7 <- read_rds("data/geocoded_results_3500.rds")
data8 <- read_rds("data/geocoded_results_4000.rds")
data9 <- read_rds("data/geocoded_results_4500.rds")
data10 <-read_rds("data/geocoded_results_5000.rds")
data11 <-read_rds("data/geocoded_results_5500.rds")
data12 <-read_rds("data/geocoded_results_6000.rds")
data13 <-read_rds("data/geocoded_results_6410.rds")

#join geocoded datasets 
shootings_w_geoid <- rbind(data1,data2,data3,data4,data5,data6,data7,data8,data9,data10,data11,data12,data13)

#pulling info from census
census_data <- get_acs(geography = "tract",
                       variables = c(population = "B01001_001", 
                                     median.gross.rent = "B25064_001",
                                     median.household.income = "B19013_001",
                                     rent.burden = "B25071_001",
                                     white = "B03002_003",
                                     af.am = "B03002_004",
                                     hispanic = "B03002_012",
                                     am.ind = "B03002_005",
                                     asian = "B03002_006",
                                     nh.pi = "B03002_007",
                                     multiple = "B03002_009",
                                     other = "B03002_008"),
                       state = c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL","IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY"), 
                       year = 2019)

#PIVOT
census_data <- pivot_wider(census_data, names_from = variable, names_sep = ".", values_from = c(estimate, moe)) 

#rename column
census_data <- rename(census_data, census_tracts_geoid = GEOID)

#join
shootings_w_acs <- shootings_w_geoid  %>%
  left_join(census_data, by=c("census_tracts_geoid"))

#updated from shreya lol
new_shootings_w_acs <- read_rds("data/new_shootings_w_acs.rds")

#adding columns
new_shootings_w_acs <- new_shootings_w_acs %>%
  mutate(tract_majority_black = case_when((estimate.af.am/estimate.population > .5) ~ "y",
                                          (estimate.af.am/estimate.population <= .5) ~ "n")) %>%
  mutate(tract_majority_white = case_when((estimate.white/estimate.population > .5) ~ "y",
                                          (estimate.white/estimate.population <= .5) ~ "n")) %>%
  mutate(tract_majority_poor = case_when((estimate.poverty_level/estimate.population > .5) ~ "y",
                                         (estimate.poverty_level/estimate.population <= .5) ~ "n")) %>%
  mutate(tract_low_income = case_when((estimate.median.household.income > 48500) ~ "y",
                                         (estimate.median.household.income <= 48500) ~ "n"))

#write file
write_rds(new_shootings_w_acs, "data/new_shootings_w_acs.rds")

#updated from shreya pt 2 lol
grouped_shootings_w_acs <- read_rds("data/grouped_shootings_w_acs.rds")
```

#Brief Analysis of Loaded Data
In looking at the police shootings data, everything looks relatively clean, so we don't foresee having to do any cleaning. We're choosing to focus on the geography of police shootings, so we need to use Census data in order to figure out where exactly these police shootings are happening, and if there are certain neighborhoods (ex: majority-minority or low income) that are more affected by police shootings. We'll likely need to form some kind of relationship between the lat/long coordinates in this dataset and data from Census tracts.

### Race Analysis

``` {r}
## first looking at data for highest percentage african american tracts: 
black_neighborhood_shootings <- new_shootings_w_acs %>%
  mutate(
    pct_black = (estimate.af.am/estimate.population) * 100)  %>%
    filter(pct_black > 50) %>%
  select(name, age, date, race, city, state, NAME, pct_black) %>%
  arrange(desc(pct_black))

## 682 shootings in neighborhoods that are majority black in a dataset of 6409

(682/ 6410) * 100
## interesting --> 10.63 percent of the total shootings were in majority black neighborhoods. 

## how many of those were black people? 
black_neighborhood_shootings %>%
  filter(race == "B") %>%
  summarize (
    count = n() 
  ) %>%
  as_tibble()

## 508 of those are black 
(508 / 682) * 100 
  = 74.4868 percent of the shootings in Black neighborhoods

## where were the most shootings? 
    #tract level
black_neighborhood_shootings %>%
  group_by(NAME) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
   ## Duval County, Florida --> Jacksonville

    #city level
black_neighborhood_shootings %>%
  group_by(city) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
  ## on the city level, most shootings are in Chicago, and many of them were in major cities 

## racial breakdown of all of the shootings
shootings_w_acs %>%
   group_by(race) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
  ##overall, white people made up more of the total shootings. 

## breakdown in majority black neighborhoods
black_neighborhood_shootings %>%
  group_by(race) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
  ##black people made up the majority
  
## white people shot in majority white neighborhoods
white_neighborhood_shootings_white <- new_shootings_w_acs %>%
filter(race == "W", estimate.white > estimate.af.am + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)
white_neighborhood_shootings_white
  ##2573 people

## black people shot in majority white neighborhoods
white_neighborhood_shootings_black <- new_shootings_w_acs %>%
filter(race == "B", estimate.white > estimate.af.am + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)
white_neighborhood_shootings_black
  ##709 people

## total people shot in majority white neighborhoods
majority_white_neighborhood_shootings <- new_shootings_w_acs %>%
filter(estimate.white > estimate.af.am + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)
majority_white_neighborhood_shootings
  ##4856 people

2573/4856 = 0.52986*100 
    = 52.986% of people shot in majority white neighborhoods are white. 

709/4856 = 0.1460049*100
    = 14.6% of people shot in majority white neighborhoods are black. 

## black people shot in majority black neighborhoods
black_neighborhood_shootings_black <- new_shootings_w_acs %>%
filter(race == "B", estimate.af.am > estimate.white + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)
black_neighborhood_shootings_black
  ##663 people

## white people shot in majority black neighborhoods
black_neighborhood_shootings_white <- new_shootings_w_acs %>%
filter(race == "W", estimate.af.am > estimate.white + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)
black_neighborhood_shootings_white
  ##144 people

## total people shot in majority black neighborhoods
majority_black_neighborhood_shootings <- new_shootings_w_acs %>%
filter(estimate.af.am > estimate.white + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)
majority_black_neighborhood_shootings
  ##1029 people

144/1029 = 0.1399417*100 
    = 13.4% of people shot in majority black neighborhoods are white. 

663/1029 = 0.6443149*100
    = 64.4% of people shot in majority black neighborhoods are black. 

```
### Income Analysis

``` {r}
## income breakdown !! 
    ## Pew Research Center defines low-income households as those earning below $48,500 
    ## Pew Research Center defines high-income households as those earning above $145,500
    
## general low-income breakdown first:
low_income_shootings <- new_shootings_w_acs %>%
  filter(estimate.median.household.income < 48500)

low_income_shootings %>% 
   summarize(
   count = n() 
  ) 
  ## 2924 shootings out of 6409 were in low-income neighborhoods 

(2924/6409) * 100
  = 45.6% of shootings were in low-income neighborhoods.

## where are the specific places:
low_income_shootings %>%
  group_by(NAME) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
  ##none of these seem to be significant

## white people shot in majority low-income neighborhoods
low_income_neighborhood_shootings_white <- new_shootings_w_acs %>%
filter(race == "W", estimate.median.household.income < 48500)
low_income_neighborhood_shootings_white
  ##1108 people

## black people shot in majority low-income neighborhoods
low_income_neighborhood_shootings_black <- new_shootings_w_acs %>%
filter(race == "B", estimate.median.household.income < 48500)
low_income_neighborhood_shootings_black
  ##865 people

1108/2924 = 0.378933*100 
    = 37.9% of people shot in majority low-income neighborhoods are white. 

865/2924 = 0.2958276*100
    = 29.6% of people shot in majority low-income neighborhoods are black. 

## general high-income breakdown
high_income_shootings <- new_shootings_w_acs %>%
  filter(estimate.median.household.income > 145500)

high_income_shootings %>% 
   summarize(
   count = n() 
  ) 
  ## 59 shootings out of 6409 were in high-income neighborhoods 

(59/6409) * 100
  = 0.92% of shootings were in high-income neighborhoods.

## where are the specific places:
high_income_shootings %>%
  group_by(NAME) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
  ##none of these seem to be significant, except maybe top 3 in Texas?

## white people shot in majority high-income neighborhoods
high_income_neighborhood_shootings_white <- new_shootings_w_acs %>%
filter(race == "W", estimate.median.household.income > 145500)
high_income_neighborhood_shootings_white
  ##33 people

## black people shot in majority high-income neighborhoods
high_income_neighborhood_shootings_black <- new_shootings_w_acs %>%
filter(race == "B", estimate.median.household.income > 145500)
high_income_neighborhood_shootings_black
  ##13 people

33/59 = 0.378933*100 
    = 55.9% of people shot in majority high-income neighborhoods are white. 

13/59 = 0.2958276*100
    = 22.0% of people shot in majority high-income neighborhoods are black. 

```
### Additional Analysis from Sean

```{r}
# All low_income and majority white tracts  
#1. What percentage of shootings happen in tracts that are low_income AND majority white?
new_shootings_w_acs %>%
  filter(tract_low_income == "y" & tract_majority_white == "y") %>%
  summarize(count = n())
  #2391 
2391/6409 = .3730691*100
    = 27.3% of shootings happen in tracts that are low_income AND majority white.

#2. What percentage of all U.S. census tracts are low_income and majority white?
grouped_shootings_w_acs %>%
  filter(tract_low_income == "y" & tract_majority_white == "y") %>%
  summarize(count = n())
  #37842 
37842/73056 = .5179862*100
    = 51.8% of all U.S. census tracts are low income and majority white.

#3. How does the answer to number 1 compare to the answer in number 2?
There is a higher percentage of U.S. census tracts that are low income and majority white than there are shootings that happen in those areas
  
# Using poverty

## All majority poor tracts
#1. What percentage of shootings happen in tracts where the majority of the population is below the poverty line? 
new_shootings_w_acs %>%
  filter(shooting_y_n == "y" & tract_majority_poor == "y") %>%
  summarize(count = n())
  #90 
90/6409 = .01404275*100
    = 1.4% of shootings happen in tracts where the majority of people are below the poverty line.

#2. What percentage of all U.S. census tracts have a majority of the population below the poverty line?
grouped_shootings_w_acs %>%
  filter(tract_majority_poor == "y") %>%
  summarize(count = n())
  #745 
745/73056 = .01019766*100
    = 1.0% of all U.S. census tracts are low income and majority white.

#3. How does the answer to number 1 compare to the answer in number 2?
```
## Post-George Floyd death vs pre-

``` {r}
new_shootings_w_acs %>%
  filter(date > "2020-05-22") %>%
  summarize(count = n())
  #1322 shootings after May 22, 2020

new_shootings_w_acs %>%
  filter(race == "B" & date > "2020-05-22") %>%
  summarize(count = n())
  #239 black people shot after May 22, 2020

  239/1322 = 18.1%
  
new_shootings_w_acs %>%
  filter(date <= "2020-05-22") %>%
  summarize(count = n())
  #5087 shootings before May 22, 2020

new_shootings_w_acs %>%
  filter(race == "B" & date <= "2020-05-22") %>%
  summarize(count = n())
  #1250 black people shot before May 22, 2020
  
  1250/5087 = 24.6%
  Percentage of Black people shot in police shootings decreased by 6.5% after the death of George Floyd.
  
new_shootings_w_acs %>%
  filter(date > "2020-05-22" & tract_majority_black == "y") %>%
  summarize(count = n())
  #125 shootings after May 22, 2020 in majority black neighborhoods

new_shootings_w_acs %>%
  filter(race == "B" & date > "2020-05-22" & tract_majority_black == "y") %>%
  summarize(count = n())
  #69 black people shot after May 22, 2020 in majority black neighborhoods

  69/125 = 55.2%
  
new_shootings_w_acs %>%
  filter(date <= "2020-05-22" & tract_majority_black == "y") %>%
  summarize(count = n())
  #557 shootings before May 22, 2020 in majority black neighborhoods

new_shootings_w_acs %>%
  filter(race == "B" & date <= "2020-05-22" & tract_majority_black == "y") %>%
  summarize(count = n())
  #440 black people shot before May 22, 2020 in majority black neighborhoods
  
  440/557 = 79.0%
  Percentage of Black people shot in police shootings in majority Black neighborhoods decreased by 23.8% after the death of George Floyd.

```

## "Armed with Toy" Analysis

```{r}
police_shootings_by_race <- new_shootings_w_acs %>%
  group_by(race) %>%
  summarize(
    total_per_race = n()
  ) 
    
police_shootings_armed_with_toy <- new_shootings_w_acs %>%
  filter(str_detect(armed, "toy")) %>%
  group_by(race) %>%
  summarize(
    total_armed_with_toy = n()
  ) 

police_shootings_armed_with_toy_per_race <- police_shootings_by_race %>%
  left_join(police_shootings_armed_with_toy) %>%
  mutate(total_armed_with_toy = replace_na(total_armed_with_toy,0)) %>%
  mutate(percent_armed_with_toy_per_race = total_armed_with_toy/total_per_race*100) %>%
  arrange(desc(percent_armed_with_toy_per_race))

new_shootings_w_acs %>%
  filter(tract_majority_black == "y" & str_detect(armed, "toy"))
  #16 toy shootings in black neighborhoods
  = 2.35% of shootings in Black neighborhoods involve the victim being armed with a toy. 

new_shootings_w_acs %>%
  filter(tract_majority_white == "y" & str_detect(armed, "toy"))
  #132 toy  shooting in white neighborhoods
  = 3.67% of shootings in white neighborhoods involve the victim being armed with a toy.
```

## "Signs of Mental Illness" Analysis

```{r}
police_shootings_by_race <- new_shootings_w_acs %>%
  group_by(race) %>%
  summarize(
    total_per_race = n()
  ) 
    
police_shootings_w_mental_illness <- new_shootings_w_acs %>%
  filter(signs_of_mental_illness == "TRUE") %>%
  group_by(race) %>%
  summarize(
    total_with_mental_illness = n()
  ) 

police_shootings_w_mental_illness_per_race <- police_shootings_by_race %>%
  left_join(police_shootings_w_mental_illness) %>%
  mutate(total_with_mental_illness = replace_na(total_with_mental_illness,0)) %>%
  mutate(percent_with_mental_illness_per_race = total_with_mental_illness/total_per_race*100) %>%
  arrange(desc(percent_with_mental_illness_per_race))

police_shootings_w_mental_illness_per_race

new_shootings_w_acs %>%
  filter(tract_majority_black == "y" & signs_of_mental_illness == "TRUE")
  #102 toy shootings in black neighborhoods
  = 15.0% of shootings in Black neighborhoods involve the victim having mental illness.

new_shootings_w_acs %>%
  filter(tract_majority_white == "y" & signs_of_mental_illness == "TRUE")
  #922 toy  shooting in white neighborhoods
  = 26.0% of shootings in white neighborhoods involve the victim having mental illness.

```

## Gender in black vs white neighborhoods Analysis

``` {r}
police_shootings_by_gender <- new_shootings_w_acs %>%
  group_by(gender) %>%
  summarize(
    total_per_gender = n()
  ) 

police_shootings_by_gender
  
new_shootings_w_acs %>%
  filter(tract_majority_black == "y" & gender == "F")
  #26 women shot in majority black neighborhoods
  26/292 = 8.9% of women are shot in majority black neighborhoods

new_shootings_w_acs %>%
  filter(tract_majority_white == "y" & gender == "F")
  #168 women shot in majority white neighborhoods
  168/292 = 57.5% of women are shot in majority white neighborhoods

new_shootings_w_acs %>%
  filter(tract_majority_black == "y" & gender == "M")
  #653 Men shot in majority black neighborhoods
  653/6113 = 10.7% of men are shot in majority black neighborhoods

new_shootings_w_acs %>%
  filter(tract_majority_white == "y" & gender == "M")
  #3424 Men shot in majority white neighborhoods
  3424/6113 = 56.0% of men are shot in majority white neighborhoods

grouped_shootings_w_acs %>%
  filter(tract_majority_black == "y")
  #6034/73056 = 8.3% of tracts are majority Black

grouped_shootings_w_acs %>%
  filter(tract_majority_white == "y")
  #49024/73056 = 67.1% of tracts are majority White

10.7% of men are shot in majority black neighborhoods, but only 8.3% of tracts are majority black 
  .107/.083 = the rate of men getting shot in majority black neighborhoods is 1.289

only 56% of men are shot in majority white neighborhoods, but 67.1% % of tracts are majority white 
  .56/.671 = the rate of men getting shot in majority white neighborhoods is 0.835

  1.289/0.835 = there are 1.54 men shot in black tracts for every 1 man shot in white tracts.
 
10.7% of men are shot in majority black neighborhoods, and 56% of men are shot in majority white neighborhoods.
  .107/.56 = the rate of men getting shot in majority black tracts to majority white tracts is 0.191 to 1.
  
8.3% of tracts are majority black, and 67.1% of tracts are majority white     
  .083/.671 = the rate of majority black tracts to majority white tracts is 0.124 to 1.
  
  .191/.124 = proportionally, there are 1.54 men shot in black tracts for every 1 man shot in white tracts.
  
new_shootings_w_acs <- new_shootings_w_acs %>%
  mutate(pct_black = (estimate.af.am/estimate.population*100)) %>%
  mutate(pct_white = (estimate.white/estimate.population*100)) 

black_male_new_shootings_w_acs <- new_shootings_w_acs %>%
  filter(race == "B" & gender == "M") %>%
  group_by(NAME) %>%
  summarize(
    pct_black,
    pct_white,
    count = n()
  )

white_male_new_shootings_w_acs <- new_shootings_w_acs %>%
  filter(race == "W" & gender == "M") %>%
  group_by(NAME) %>%
  summarize(
    pct_black,
    pct_white,
    count = n()
  )

#white male with pct_white
white_male_new_shootings_w_acs %>%
  ggplot() +
  geom_point(aes(x=pct_white,y=count)) +
  geom_smooth(aes(x=pct_white,y=count), method="lm")

#white male with pct_black
white_male_new_shootings_w_acs %>%
  ggplot() +
  geom_point(aes(x=pct_black,y=count)) +
  geom_smooth(aes(x=pct_black,y=count), method="lm")

#black male with pct_black
black_male_new_shootings_w_acs %>%
  ggplot() +
  geom_point(aes(x=pct_black,y=count)) +
  geom_smooth(aes(x=pct_black,y=count), method="lm")

#black male with pct_white
black_male_new_shootings_w_acs %>%
  ggplot() +
  geom_point(aes(x=pct_white,y=count)) +
  geom_smooth(aes(x=pct_white,y=count), method="lm")

```

## National Per Capita Analysis

``` {r}
grouped_shootings_w_acs %>%
  mutate(total_white = sum(grouped_shootings_w_acs$estimate.white)) %>%
    #197,100,373 total white Americans
  mutate(total_black = sum(grouped_shootings_w_acs$estimate.af.am)) %>%
    #39,977,554 total Black Americans
  mutate(total_AAPI = sum(grouped_shootings_w_acs$estimate.asian + grouped_shootings_w_acs$estimate.nh.pi)) %>%
    #18,249,465 total AAPI
  mutate(total_indigenous = sum(grouped_shootings_w_acs$estimate.am.ind)) %>%
    #2,160,378 total Indigenous Americans
  mutate(total_hispanic = sum(grouped_shootings_w_acs$estimate.hispanic)) %>%
    #58,479,370 total Hispanic Americans
  mutate(total_other = sum(grouped_shootings_w_acs$estimate.other + grouped_shootings_w_acs$estimate.multiple)) %>%
    #8,730,655 total "other" or mixed Americans
  
  grouped_shootings_w_acs

national_grouped_shootings_w_acs
  
  
police_shootings_by_race





```

## State Per Capita Analysis

``` {r}


```
