---
title: "Data Analysis Project"
author: "michael purdie"
date: "11/14/21"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are working with the [Washington Post police shooting database](https://github.com/washingtonpost/data-police-shootings) and the [Washington Post homicides database.](https://github.com/washingtonpost/data-homicides)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
#essential libraries
library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
#potential libraries 
library(janitor)
library(lubridate)

#API Census key
census_api_key("b559c55dc9a86baaa9ab42174d2fa978a618dff1")

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
#Load required data
police_shootings <- read_csv("data/data-police-shootings-master/fatal-police-shootings-data.csv")

#cleaning and organizing
#geocoded data
data1 <- read_rds("data/geocoded_results_500.rds")
data2 <- read_rds("data/geocoded_results_1000.rds")
data3 <- read_rds("data/geocoded_results_1500.rds")
data4 <- read_rds("data/geocoded_results_2000.rds")
data5 <- read_rds("data/geocoded_results_2500.rds")
data6 <- read_rds("data/geocoded_results_3000.rds")
data7 <- read_rds("data/geocoded_results_3500.rds")
data8 <- read_rds("data/geocoded_results_4000.rds")
data9 <- read_rds("data/geocoded_results_4500.rds")
data10 <-read_rds("data/geocoded_results_5000.rds")
data11 <-read_rds("data/geocoded_results_5500.rds")
data12 <-read_rds("data/geocoded_results_6000.rds")
data13 <-read_rds("data/geocoded_results_6410.rds")

#join geocoded datasets 
shootings_w_geoid <- rbind(data1,data2,data3,data4,data5,data6,data7,data8,data9,data10,data11,data12,data13)

#pulling info from census
census_data <- get_acs(geography = "tract",
                       variables = c(population = "B01001_001", 
                                     median.gross.rent = "B25064_001",
                                     median.household.income = "B19013_001",
                                     rent.burden = "B25071_001",
                                     white = "B03002_003",
                                     af.am = "B03002_004",
                                     hispanic = "B03002_012",
                                     am.ind = "B03002_005",
                                     asian = "B03002_006",
                                     nh.pi = "B03002_007",
                                     multiple = "B03002_009",
                                     other = "B03002_008"),
                       state = c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL","IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY"), 
                       year = 2019)

#PIVOT
census_data <- pivot_wider(census_data, names_from = variable, names_sep = ".", values_from = c(estimate, moe)) 

#rename column
census_data <- rename(census_data, census_tracts_geoid = GEOID)

#join
shootings_w_acs <- shootings_w_geoid  %>%
  left_join(census_data, by=c("census_tracts_geoid"))


```

#Brief Analysis of Loaded Data
In looking at the police shootings data, everything looks relatively clean, so we don't foresee having to do any cleaning. We're choosing to focus on the geography of police shootings, so we need to use Census data in order to figure out where exactly these police shootings are happening, and if there are certain neighborhoods (ex: majority-minority or low income) that are more affected by police shootings. We'll likely need to form some kind of relationship between the lat/long coordinates in this dataset and data from Census tracts.

### Analysis

``` {r}
## first looking at data for highest percentage african american tracts: 
black_neighborhood_shootings <- shootings_w_acs %>%
  mutate(
    pct_black = (estimate.af.am/estimate.population) * 100)  %>%
    filter(pct_black > 50) %>%
  select(name, age, date, race, city, state, NAME, pct_black) %>%
  arrange(desc(pct_black))

## 682 shootings in neighborhoods that are majority black in a dataset of 6409

(682/ 6410) * 100
## interesting --> 10.63 percent of the total shootings were in majority black neighborhoods. 

## how many of those were black people? 
black_neighborhood_shootings %>%
  filter(race == "B") %>%
  summarize (
    count = n() 
  ) %>%
  as_tibble()

## 420 of those are black 
(509 / 682) * 100
## which is 74.63 percent of the shootings in Black neighborhoods

## where were the most shootings? 
    #tract level
black_neighborhood_shootings %>%
  group_by(NAME) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
   ## Duval County, Florida --> Jacksonville

    #city level
black_neighborhood_shootings %>%
  group_by(city) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))

  ## on the city level, most shootings are in Chicago, and many of them were in major cities 

## racial breakdown of all of the shootings
shootings_w_acs %>%
   group_by(race) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
## overall, white people made up more of the shootings. 

## breakdown in majority black neighborhoods
black_neighborhood_shootings %>%
  group_by(race) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))

## income breakdown !! 
    ## Pew Research Center defines low income households as those earning below $48,500 
    
## general breakdown first:
low_income_shootings <- shootings_w_acs %>%
  filter(estimate.median.household.income < 48500) %>% 
   summarize(
   count = n() 
  ) 
## 2924 shootings out of 6409 were in low-income neighborhoods 
(2924/6409) * 100
## that's 45 percent !! 

## where are the specific places:
low_income_shootings %>%
  group_by(NAME) %>%
   summarize(
   count = n() 
  ) %>%
  arrange(desc(count))
  
## white people shot in majority white neighborhoods
white_neighborhood_shootings_white <- shootings_w_acs %>%
filter(race == "W", estimate.white > estimate.af.am + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)

## white people shot in majority white neighborhoods
white_neighborhood_shootings_black <- shootings_w_acs %>%
filter(race == "B", estimate.white > estimate.af.am + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)

## total people shot in majority white neighborhoods
majority_white_neighborhood_shootings <- shootings_w_acs %>%
filter(estimate.white > estimate.af.am + estimate.am.ind + estimate.asian + estimate.nh.pi + estimate.other)

1887/3599 = 0.5243123 ##will be different with correct data
= 52.5% of people shot in majority white neighborhoods are white.

527/3599 = 0.1464296 ##will be different with correct data
= 14.6% of people shot in majority white neighborhoods are black.
```

### Question 3

* **Question text**: Are there patterns regarding race and instances in which the victim was armed with a toy weapon? 

* **Analysis summary**: 

```{r}
police_shootings_by_race <- police_shootings %>%
  group_by(race) %>%
  summarize(
    total_per_race = n()
  ) 
    
police_shootings_armed_with_toy <- police_shootings %>%
  filter(str_detect(armed, "toy")) %>%
  group_by(race) %>%
  summarize(
    total_armed_with_toy = n()
  ) 

police_shootings_armed_with_toy_per_race <- police_shootings_by_race %>%
  left_join(police_shootings_armed_with_toy) %>%
  mutate(total_armed_with_toy = replace_na(total_armed_with_toy,0)) %>%
  mutate(percent_armed_with_toy_per_race = total_armed_with_toy/total_per_race*100) %>%
  arrange(desc(percent_armed_with_toy_per_race))

police_shootings_by_race
police_shootings_armed_with_toy
police_shootings_armed_with_toy_per_race

#not finished, trying to figure out if i need to left_join bc ungroup() wouldnt work
```