---
title: "Data Analysis Project"
author: "michael purdie"
date: "11/14/21"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are working with the [Washington Post police shooting database](https://github.com/washingtonpost/data-police-shootings) and the [Washington Post homicides database.](https://github.com/washingtonpost/data-homicides)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
#essential libraries
library(tidyverse)
library(janitor)
library(lubridate)
#additional libraries 
library(sf)
library(tigris)
library(tidycensus)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
police_shootings <- read_csv("data/data-police-shootings-master/fatal-police-shootings-data.csv")
homicides <- read_csv("data/data-homicides-master/homicide-data.csv")

```