---
title: "lab_02"
author: "michael purdie"
date: "9/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse and Lubridate. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and lubridate
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with West Virginia loans contained in the PPP loans data.  

**Task** Run the codeblock below to load the data.

```{r}
west_virginia_ppp <- read_rds("lab_02.rds")
```

## Examine Data

## Answer questions

**Q1** Create a new column that divides the loan amounts into multiple options: under_10k, 10k_50k, 50k_100k, 100k_500k, 500k_1m, 1m_plus. How many loans are in each group? You will need to use `case_when` to make this work and there's an example of this in the book chapter on mutate.
**A1** There are 9,310 loans less than $10,000, 15,372 loans between $10,000 and $50,000, 3,978 loans between $50,000 and $100,000, 4,152 loans between $100,000 and $500,000, 540 loans between $500,000 and $1,000,000, and 356 more than $1,000,000.

```{r}
west_virginia_ppp %>%
  mutate(
    loan_brackets = case_when(
      amount < 10000 ~ 'under_10k',
      amount >= 10000 & amount < 50000  ~ '10k_50k', 
      amount >= 50000 & amount < 100000  ~ '50k_100k',
      amount >= 100000 & amount < 500000 ~ '100k_500k',
      amount >= 500000 & amount < 1000000 ~ '500k_1m',
      amount > 1000000  ~ '1m_plus'
    ) 
  ) %>%
  group_by(loan_brackets) %>%
  summarize(
    count_loans = n()
  )
```

**Q2** Which servicing lenders have the most loans with undisbursed amounts greater than 0?
**A2** Harvest Small Business Finance, LLC has the most loans with undisbursed amounts greater than 0 at 139 loans.

```{r}
west_virginia_ppp %>%
  mutate(
    undisbursed_greater_than_0 = if_else(
      undisbursed_amount > 0, "Yes", "No"
    )
  ) %>% filter(undisbursed_greater_than_0 == 'Yes') %>%
  group_by(servicing_lender_name) %>%
  summarize(
    count_undisbursed_loans = n()
  ) %>%
  arrange(desc(count_undisbursed_loans))

```

**Q3** How would you describe the top 10 list of those lenders? What do they have in common or not? (You may want to look up some of the top ones.)
**A3** Many of these are Limited Liability Companies (LLCs), as well as work with mortgages and finances.

**Q4** Which month in 2021 had the most loans approved?
**A4** March 2021, with 4,404 loans

```{r}
west_virginia_ppp %>%
  filter(date_approved > '2020-12-31') %>%
  mutate(Month = floor_date(date_approved, "month")) %>% 
  group_by(Month) %>%
  summarise(total = n()) %>%
  arrange(desc(total))

```
