---
title: "Reverse Engineering Project"
author: "michael purdie"
date: "10/10/21"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Over a Dozen Black and Latino Men Accused a Cop of Humiliating, Invasive Strip Searches. The NYPD Kept Promoting Him.](https://www.propublica.org/article/over-a-dozen-black-and-latino-men-accused-a-cop-of-humiliating-invasive-strip-searches-the-nypd-kept-promoting-him)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
#install readxl
install.packages("readxl")

#load tidyverse
library(tidyverse)
#load janitor
library(janitor)
#load lubridate
library(lubridate)
#load readxl
library(readxl)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
allegations <- read_csv("data/CCRB-Complaint-Data_202007271729/allegations_202007271729.csv")
allegation_description <- read_excel("data/CCRB-Complaint-Data_202007271729/CCRB Data Layout Table.xlsx")

#cleaning data
cleaned_allegations <- allegations %>%
  #new cleaned board_disposition column
  mutate(
    cleaned_board_disposition = case_when(
      str_detect(board_disposition, "^S") ~ "Substantiated",
      TRUE ~ board_disposition
    )
  ) %>%
  #new column that places officers into "yes" or "no
  mutate(
    rank_above_captain = if_else(
      rank_now == 'Deputy Inspector' | 
      rank_now == 'Inspector' | 
      rank_now == 'Deputy Chief' | 
      rank_now == 'Assistant Chief' | 
      rank_now == 'Bureau Chief' | 
      rank_now == 'Police Comissioner' | 
      rank_now == 'Chiefs and other ranks', "Yes", "No")
  )


```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “A dozen of those high-ranking officers have three or more complaints with substantiated allegations; 10 are white. Five have reached the rank of deputy inspector, the lowest promoted at the commissioner’s discretion; all are white.”

* **Analysis summary**: My analysis did not confirm the results from the article, but I believe this is due to the fact that the data has been updated since. I found that there were 25 high-ranking officers with three or more complaints, not a dozen, and 17 were White, not 10. Also, there were 11 high-ranking officers with three or more complaints that are now a deputy inspector, and they are not all White. In regard to format, I narrowed down the data to the appropriate requirements, then grouped and summarized to obtain the number of total complaints for each officer. Finally, I filtered for officers with more than two complaints.

```{r}
#High-ranking officers with three or more complaints
cleaned_allegations %>%
  filter(rank_above_captain == "Yes") %>%
  filter(cleaned_board_disposition == "Substantiated") %>%
  group_by(unique_mos_id, first_name, last_name, rank_now, mos_ethnicity) %>%
  summarize(
    count = n()
  ) %>%
    filter(count > 2) %>%
    arrange(desc(count))

#White high-ranking officers with three or more complaints
cleaned_allegations %>%
  filter(rank_above_captain == "Yes") %>%
  filter(cleaned_board_disposition == "Substantiated") %>%
  filter(mos_ethnicity == "White") %>%
  group_by(unique_mos_id, first_name, last_name, rank_now, mos_ethnicity) %>%
    summarize(
      count = n()
    ) %>%
      filter(count > 2) %>%
      arrange(desc(count))


#High-ranking officers with three or more complaints that have reached the rank of deputy inspector
cleaned_allegations %>%
  filter(rank_now == "Deputy Inspector") %>%
  filter(cleaned_board_disposition == "Substantiated") %>%
  group_by(unique_mos_id, first_name, last_name, rank_now, mos_ethnicity) %>%
    summarize(
      count = n()
    ) %>%
      filter(count > 2) %>%
      arrange(desc(count))

```

### Sentence 2

* **Sentence text**: “Eighty-six of the roughly 420 officers in the department who currently hold a rank above captain — running precincts and other large commands and overseeing hundreds of officers — have tallied at least one misconduct allegation that was substantiated by the CCRB, meaning that investigators amassed enough evidence of offenses, ranging from bad language to pistol whippings, to say that they happened and broke patrol guidelines. The most common involved improper property entries and searches.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: “Of at least 77 allegations made against him in 26 separate CCRB complaints, 29 were unsubstantiated; five were “unfounded,” meaning investigators concluded the incident never took place; and 27 were “exonerated,” meaning the conduct fell within patrol guidelines.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: “But many of the officers also said that it’s possible to be an effective cop without so many complaints. An overwhelming majority of the city’s 36,000 officers do not have a single substantiated complaint on their records; only 1 in 9 does.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
#make new dataset that gives number of officers with at least one substantiated complaints
officers_with_complaints <- cleaned_allegations %>%
  filter(cleaned_board_disposition == "Substantiated") %>%
  distinct(unique_mos_id, first_name, last_name, rank_now)

#number of officers with at least one substantiated complaint/total number of officers
officers_with_complaints %>%
    summarize(
      total = n()/36000
    ) 

#1 in 9 = .1111

```

### Sentence 5

* **Sentence text**: There isn't really a sentence from the story, but I'd like you to answer this question: The story focused on Christopher McCormack. Use data analysis to justify that decision. Why does he stand out as newsworthy, from all of the people they could have selected? Why, do you think, did the reporters single him out for an in-depth story. Write code to support your argument.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-