---
title: "lab_11"
author: "michael purdie"
date: "11/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. You'll need tidyverse, tidycensus, janitor, sf and tigris, and you'll also need to install your census API key.
``` {r}
#load packages
library(tidyverse)
library(sf)
library(janitor)
library(tigris)
library(tidycensus)
library(lubridate)

#API Census key
census_api_key("b559c55dc9a86baaa9ab42174d2fa978a618dff1")
```
## Load data

**Task** Create a codeblock and load the WV slice of the PPP loan data, which is in the data folder and is called lab_11.rds.

``` {r}
#load data
wv_ppp <- read_rds("data/lab_11.rds")
```

## Questions 

**Q1.** You are interested in doing some reporting on how West Virginia companies in the "Mining, Quarrying, and Oil and Gas Extraction" industry (https://www.naics.com/six-digit-naics/?code=21) used PPP loans during the pandemic (West Virginia is a big coal mining state).  

Then produce a choropleth map that shows West Virginia counties shaded according to the total amount of loans to that industry per 100,000 population. 

Use the magma color palette, and shade on a "log" scale. Also produce a table that shows the total amount of loans per 100,000 population for each county, ranked from highest to lowest, along with the total amount of loans and the total number of loans. 

Then answer these questions:

* Based on the map, what region of the state appears to have gotten more loans in this industry? What geographic pattern do you notice in the data?
* Based on the table, what county got the highest total loan amount? What was the total amount per 100K population, the total amount of loans (not population adjusted), how many loans were there, and what's the population?

**A1.**  According to the map, Richie County got the highest total amount of loans per 1000k population with $77,752,742.79, and according to the table, the highest total loan amount at $7,653,980.00, with a population of 9,844, and a total of 9 mining PPP loans. More mining in the Appalachian Mountains, and less people live there.

``` {r}
#create dataset that isolates West Virginia mining PPP loans
wv_mining_ppp <- wv_ppp %>%
  filter(str_detect(naics_code, "^21")) %>%
  #get total amount of loans in dollar amount per county
  group_by(project_county_name) %>%
  summarize(
    total_mining_ppp_per_county = sum(amount),
    total_mining_ppp_loans_per_county = n()
  )

#using Census Key, pull WV county populations and geometry
wv_county_population <- get_acs(geography = "county", 
              variables = c(population = "B01001_001"), 
              state = "WV",
              geometry = TRUE)

#rename columns in wv_county_population dataset and clean
wv_county_population <- wv_county_population %>%
  rename(project_county_name = NAME) %>%
  mutate(project_county_name = toupper(str_remove_all(project_county_name,", West Virginia|County"))) %>%
  mutate(project_county_name = str_trim(project_county_name,side="both")) %>%
  rename(population = estimate) %>%
  select(project_county_name, population, geometry)

#join dataframes together
wv_mining_ppp_per_100k <- wv_county_population %>%
  left_join(wv_mining_ppp) %>%
  #make tibble
  as_tibble() %>%
  #change NAs
  mutate(total_mining_ppp_per_county = replace_na(total_mining_ppp_per_county,0)) %>%
  mutate(total_mining_ppp_loans_per_county = replace_na(total_mining_ppp_loans_per_county,0)) %>%
  #do math
  mutate(mining_ppp_per_100k = total_mining_ppp_per_county/population*100000) %>%
  arrange(desc(mining_ppp_per_100k))

#display dataset
wv_mining_ppp_per_100k

#make map
ggplot() + 
  geom_sf(data=wv_mining_ppp_per_100k, aes(fill=mining_ppp_per_100k, geometry=geometry)) + 
  theme_minimal() +
  scale_fill_viridis_b(option="magma", trans = "log") 
```

**Q2.** You want to understand WHY the spatial pattern you saw above exists.  So you grab a shapefile with the latitude and longitude coordinates of every mine permit in West Virginia from the state Department of Environmental Protection. 

Here's the url: https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json. 

(It's in a slightly different format than the .shp file in the lab. It's another type of shapefile called "geojson".  It will work just the same, though, despite that difference. You can read it directly into R without manually downloading it, by putting the url in the st_read() function we learned about in this week's prelab.) 

This data includes both currently operating mines and old ones that have closed.  You want to examine only mines that have a permit expiration date after Sept. 1, 2021. 

Make a map with a point for each one of these mine permits and put a blank map with the outline of West Virginia counties (which you can get from Tigris) behind it. 

What spatial pattern do you see? How does it compare to the spatial pattern in the first map you made?

**A2.** It kind of corresponds to the orange areas in the first map, as they are concentrated in the Appalachian Mountains.
``` {r}
#read url into dataframe
wv_mines <- st_read("https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json")

#mutate and filter for after Sept. 1, 2021 on WV mines dataframe
wv_mines <- wv_mines %>%
  #make a tibble
  as_tibble() %>%
  mutate(expiredate = ymd(expiredate)) %>%
  filter(expiredate > "2021-09-01")

#pull counties geometry from tigris
counties <- counties()

#filter for only WV counties
wv_counties <- counties %>%
  filter(STATEFP == "54")

#plot for counties, then mines
ggplot() + 
  geom_sf(data=wv_counties, fill="white") + 
  geom_sf(data=wv_mines, aes(geometry=geometry)) +
  theme_minimal() 
```

**Q3.** To confirm the spatial comparison you identified in question 2, let's create a new map that layers the mine location points on top of our PPP choropleth map we created in question 1. What do you notice about the amount of mines in the county that has the highest total loan amount per 100K?

**A3.** There appears to only be one mine.
``` {r}
#plot for wv_mining_ppp_per_100k, then mines
ggplot() + 
  geom_sf(data=wv_mining_ppp_per_100k, aes(fill=mining_ppp_per_100k, geometry=geometry)) + 
  geom_sf(data=wv_mines, aes(geometry=geometry)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma", trans = "log") 
```

**Q4.** Starting from the original WV ppp dataframe, examine the loans to companies in "Mining, Quarrying, and Oil and Gas Extraction" in the county with the highest total loan amount per 100K (you can use the table you created in question 1 to identify the top county)?
* What single company accounts for the biggest chunk of the total loan amount? 
* What specific industry are they in, based on their 6-digit NAICS code? Look them up on the web. What services do they provide?
* How many jobs did they save, according to the data?
* What's your hypothesis for why a company in their specific industry would need to seek a PPP loan during the pandemic?

**A4.** Hall Drilling, LLC accounts for $68,666,000. They are in the "drilling oil and gas wells" industry, and they saved 387 jobs with the loan. This company probably needed the funds to keep paying workings during the pandemic as energy consumption went down.
``` {r}
#using WV PPP data, filtering for Ritchie County and mining industry
wv_ppp %>%
  filter(str_detect(naics_code, "^21")) %>%
  filter(project_county_name == "RITCHIE") %>%
  arrange(desc(amount))
```
-30-
