---
title: "lab_01"
author: "michael purdie"
date: "9/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**). 
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**). 
* write out the answer in the form of a complete sentence in the space given (look for **A**). 
When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS. 

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse.To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}


# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data 

We're going to load the slice of PPP loans for West Virginia.

**Task** Run the codeblock below to load the data. 

```{r}
wv_ppp_loans <- read_rds("lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data. 
**Q1** How many rows and columns does this dataframe have?
**A1** 33,703 rows, 63 columns

```{r}
glimpse(wv_ppp_loans)
```

## Answer questions

**Q2** Which West Virginia county got the most loans? How many?
**A2** Kanawha County, 4332 loans

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarize(
    count_loans = n()
  ) %>%
  arrange(desc(count_loans))
```
**Q3** Which West Virginia county got the fewest loans? How many?
**A3** Wirt County, 36 loans

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarize(
    count_loans = n()
  ) %>%
  arrange(count_loans)
```
**Q4** What's your best educated guess to explain why the county identified in question 2 had a lot of loans and why the county identified in question 3 had only a few loans? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** Kanawha County has a higher population than Wirt County, which means it has more businesses and thus more loans.


**Q5** What West Virginia business got the state's largest PPP loan? How much was it for? What does the business do? How many jobs did the loan retain, according to the data?
**A5** Greenbriar Hotel Corporation got the state's largest PPP loan of $8,912,202, which is labeled as a "corporation." The loan retained 500 jobs.

```{r}
wv_ppp_loans %>%
   arrange(desc(amount))
```

**Q6** What West Virginia politician (with his family) owns the company identified in question 2? What's your source for that information? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** I think its supposed to say question 5, in which case the answer is Gov. Jim Justice, according to propublica.org.

**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by county):
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings. 

**A7** Of the 33,703 loans given out in West Virginia, there was a wide spread of amounts. The mean loan amount was $77,338 and the median loan amount was $20,832, which means that there were multiple loan amounts that were so large that they skew the datas average. This represents a disproportionate amount of money given to different companies.
```{r}
wv_summary <- wv_ppp_loans %>%
  group_by(state) %>%  
  summarize(
   count_loans = n(),
   total_loan_amount = sum(amount),
   average_loan = mean(amount),
   median_loan = median(amount),
   smallest_loan = min(amount),
   largest_loan = max(amount)
  )
```

**Q8** Produce a summary table that has the following information for each West Virginia county:
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_county_summary` so that you can examine it with the data viewer. 

What newsworthy nuggets or patterns jump out at you that you could imagine dropping into a news story or doing follow up reporting on? Write up a few sentence that summarizes one county. 

**A8** Kanawha County received a total of $420,270,037 in loans, which is almost twice as much as any other county; however, they also recieved significantly more loans than any other county at 4332. The largest loan was in Greenbrier County.
```{r}
wv_county_summary <- wv_ppp_loans %>%
  group_by(project_county_name) %>%  
  summarize(
   count_loans = n(),
   total_loan_amount = sum(amount),
   average_loan = mean(amount),
   median_loan = median(amount),
   smallest_loan = min(amount),
   largest_loan = max(amount)
  ) 
```
